visualFields 1.0.7: Tools for Analyzing the Field of Vision in \proglang{R}

### 2025-05-15 jss-article-rnw/article.Rnw
Copyedited rest of the article.

The package is available on CRAN under the Apache 2.0 license and can be installed with the following command:

<<eval=FALSE>>=

install.packages("visualFields")

To install the newest version of the package, run:

<<eval=FALSE>>=

library(devtools)
install_github("imarinfr/vf1/source")

<<echo=FALSE>>=

library(ggplot2)
library(plotrix)

<<echo=FALSE, fig=TRUE>>=

library(visualFields)
library(plotrix)

default_locmap <- getlocmap()

plot(x = default_locmap$coord$x, 
     y = default_locmap$coord$y, 
     type = "n", 
     asp = 1,
     xlim = c(-30, 30),
     ylim = c(-34, 34),
     yaxt = 'n',
     xaxt = 'n',
     bty = 'n',
     xlab = "",
     ylab = "",
     main = "24-2 grid of test locations")
lines(-c(30, 2.5), c(0, 0), lwd = 2, col = "lightgray")
lines(c(2.5, 30), c(0, 0), lwd = 2, col = "lightgray")
lines(c(0, 0), -c(30, 2.5), lwd = 2, col = "lightgray")
lines(c(0, 0), c(2.5, 30), lwd = 2, col = "lightgray")
lines(x = c(-2.5, 2.5), y = c(0, 0), lwd = 1.5)
lines(x = c(0, 0), y = c(-2.5, 2.5), lwd = 1.5)
for(r in c(10, 20, 30))
  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
draw.ellipse(15, -1.5, 2.75, 3.75, col = "gray", border = NA)
text(x = 0, y = -34, "horizontal eccentricity [degrees]")
text(x = -34, y = -0, "vertical eccentricity [degrees]", srt = 90)
text(x = 31, y = 0, "30\u00b0", adj = c(0, 0.5))
text(x = 0, y = 31, "30\u00b0", adj = c(0.5, 0))
points(x = default_locmap$coord$x[-default_locmap$bs],
       y = default_locmap$coord$y[-default_locmap$bs])
points(x = default_locmap$coord$x[default_locmap$bs],
       y = default_locmap$coord$y[default_locmap$bs],
       col = "red", pch = 19)

\subsection{Grid patterns}
The spatial position of the points $l_1$ to $l_k$ in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter. When analyzing data, you need to specify the locmap.

By default, the 24-2 pattern (shown in the previous figure) is used, which consists of 54 locations that are equally spaced across the field.

<<>>=
str(default_locmap)
@

Beside a name and a description, a locmap dataset contains a coordinate table with the x- and y-coordinates of the locations measured in degrees, and the id of the points that cover the physiological blind spot (the id is the number x in $l_x$ and corresponds to the row in the coord table).

Other grid patterns are also defined in the package. For instance, the 10-2 grid:

@@

And many others:

@@

Locations of the locmap that cover the blind spot are reported in the locmaps lists under the variable bs. If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to `r numeric(0)`.

@@

The first columns characterize the patient, the time and duration of the examination, the diagnosis, and the reliability parameters 'false positive responses', 'false negative responses' and 'fixation losses'.

Each row corresponds to one visual field test. Visual field testing is performed monocularly (one eye at a time), so that two examinations exist for each date. The abbreviations for the laterality ('eye') is derived from latin (OD: right eye = oculus dexter, and OS: left eye = oculus sinister), when both eyes are examined, the abbreviation is OU.

The environment variable locini contains the number of the first column that contains the actual visual field date. Each column $l_{x}$ corresponds to one of {k} locations in the visual field test. A concise way to select the columns $l_1$ to $l_{k}$ is to use the \textbf{getvfcols} function.

getlocini()
getvfcols()

vfpwgSunyiu24d2[1, getvfcols()]

There mustn't be any columns after these columns. Use the \textbf{vffilter} function to select specific fields and the \textbf{vfisvalid} function to check whether a table fulfills the criteria. The latter will also tell you what the problem is when it doesn't.

example1 <- vffilter(vfpwgSunyiu24d2, 
                     id == "sample1" & 
                       eye == "OD")[1, ]

example1$newVariable <- "not valid"

vfisvalid(example1)

\subsection{Mathematical derivatives of contrast thresholds for statisitcal analysis}

\subsubsection{Identifying defective locations on the visual field}

As mentioned earlier, the sensitivity in dB at each location is calculated from the contrast threshold relative to the maximal contrast.

The sensitivity to light fluctuates considerably between measurements in the same observer (test-retest variability ---and more as the disease progresses) and between observers (inter-individual differences). They also decrease at different rates depending on the eccentricity with age.

With these predictable and unpredictable differences in the pointwise sensitivity measurements, how can we provide tools and reports to aid the clinical assessment of the visual field? Or more to the point, how can we discriminate between healthy locations and locations that are damaged or suspected of being damaged? Clinically, it is more useful to compute and report the deviations from an age-corrected mean normal visual field, since the identification of spatial patterns of damage becomes easier, thus facilitating the clinical assessment of the visual field.

But this analysis requires the use of a \textbf{normative dataset} from which to obtain models describing the age effect on sensitivity at each location. These functions are usually linear and those \textbf{age linear models} are used to obtain the \textbf{age-corrected mean normal visual field}.

The resulting \textbf{deviations} from age-corrected mean normal visual field are also measured in dB and they are interpreted as follows. By convention, in the Humphrey Field Analyzer, a negative deviation value means that the sensitivity at that location is below mean normal. In contrast, on Octopus perimeters, negative is above and positive below mean normal (therefore, values are called \textbf{defects}). In this package, those differences are called \textbf{total deviation} and their interpretation are as for the Humphrey perimeter. In case you use visualFields with data from a different perimeter it is important to know what is the definition of \textbf{deviation} and change the sign if necessary.

In eye diseases like glaucoma, general defects that affect the whole visual field are often less relevant than focal scotomas. And often ocular opacities caused by cataract can mask such focal defects. For this reason, a statistical post-hoc correction to \textbf{total deviation} values is further applied to obtain \textbf{pattern deviation} values. The correction factor is often called the \textbf{general height} and for the 24-2 Humphrey is obtained by selecting the 7th greatest total deviation value. The \textbf{pattern deviation} values equal the \textbf{total deviation} values minus the \textbf{general height}.

\subsubsection{Normative dataset and normative references}
The visualFields package contains three datasets of healthy subjects that can be used to obtain normative age linear models and references.

The SUNY-IU dataset [@IMF:Mar2013b] for the 24-2 SITA Standard and the Goldmann Size III:
<<SUNYIU datasets>>=
str(vfctrSunyiu24d2[,1:10])
@

The Iowa dataset [@IMF:Marin2018c] for a bimodal ZEST algorithm and the Goldmann Size V:
<<Iowa datasets>>=
str(vfctrIowaPC26[,1:10])
str(vfctrIowaPeri[,1:10])
@

From these datasets, or any other dataset of healthy eyes, normative references can be computed as follows:

<<nv>>=
nv <- nvgenerate(vfctrSunyiu24d2)
@

For future use, we set `nv` as the normative values to be used in visualFields for the calculation of total and pattern deviation values and other statistical analyses.
<<setnv>>=
setnv(nv)
@

\subsubsubsection{Age linear models and mean normal visual fields}
The normative values and references stored in `nv` contains the coefficients of the age linear models:

<<agelm>>=
print(nv$agem$coeff)
@

And a function from where the age-corrected mean normal visual fields can be computed:
<<agemeannormal>>=
age <- 50
print(nv$agem$model(age))
@

Notice that the locations corresponding to the anatomical region of the blind spot all have `NA` as values, as they are excluded from the statistical analysis.

\subsubsection{Total deviation}
Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable \textbf{nv} (normal values).

<<sensitivities>>=

example2 <- vffilter(vfpwgRetest24d2, 
                     id == "1" & 
                       eye == "OD")[1,]

print(example2[, getvfcols()[1:10]])

<<total deviation>>=

age <- example2$age[1]

print(age)

normal_values <- getnv()$agem$model(age)[1:10]

print(normal_values)

total_defects <- example2[, getvfcols()[1:10]] - normal_values

print(total_defects)

These values can be automatically calculated with a dedicated function.

td <- gettd(example2)
print(td[, getvfcols()[1:10]])

\subsubsection{Pattern deviation}
Pattern deviation can be calculated after obtaining the general height.
<<gh>>=
gh <- getgh(td)
print(gh)
pattern_defects <- td[, getvfcols()[1:10]] - gh
print(pattern_defects)
@

These values can be automatically calculated with a dedicated function.

<<pattern deviation>>=

pd <- getpd(td)
print(pd[, getvfcols()[1:10]])

\subsubsection{Total and Pattern deviation probabilities}
As mentioned psychophysical sensitivity values vary considerably from test to test and between subjects Therefore, even in perfectly average eyes deviations will be different from zero most of the time. Given a visual field from a patient, the clinician has to know if a given deviation is likely to be caused by disease as opposed to normal retest and inter-individual variability.

The function `nvgenerate` computes for the normative dataset, the total deviation and pattern deviation values for healthy eyes and records pointwise empirical probability distribution functions [@IMF:Hei1987b] that can be used to compute the probability that a specific total deviation (or pattern deviation value) value is from a healthy location [@IMF:Hei1987a] [@IMF:Heijl1991]. Thus, a low probability implies a greater likelihood of the location being damaged. Typically (and in our example), the probabilities are categorized as
<<probability categories>>=
print(getnv()$lut$probs)
@

where `0.005` means probability $< 0.005$, where `0.01` means probability $< 0.01$, etc. A probability greater than 0.05 is considered within the normal range.

The probability values for the previous values are as follows.

<<probability maps>>=
print(gettdp(td)[, getvfcols()[1:10]])
print(getpdp(pd)[, getvfcols()[1:10]])
@

\subsection{Global indices}
Global indices summarize the visual field in a few parameters and their values and probability level can be calculated with the visualField package. The resulting table contains the visual field header data (patient id, eye, etc.) and 8 additional columns (msens, ssens, tmd, tsd, pmd, psd, gh, and vfi).

print(getgl(example2)[,-(1:getlocini()-1)])

The column `msens` is the mean sensitivity (MS); the arithmetic mean of all sensitivity values. The column `ssens` is the standard deviation of sensitivity values.

The column `tmd` is the mean deviation (MD); the arithmetic mean of all total deviation values. The column `tsd` is the standard deviation of total deviation values.

The column `pmd` is the mean of pattern deviation values. The column `psd` is the standard pattern deviation (PSD) or the standard deviation of the pattern deviation values.

The last column, `vfi`, is the visual field index (VFI) [@IMF:Ben2008], and the second to last is the general height (GH).

The probability of the observed values given that the visual field is normal can also be calculated.

print(getglp(getgl(example2))[, -(1:getlocini()-1)])

@@

\subsection{Plotting the visual field}
There are different ways to visualize the perimetric results. These visualizations have to be interpreted like maps in that they represent a simplified version of the hill of vision.

The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot. Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values. These data are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other perimeters.

By convention, when both eyes are presented side by side, the left eye is shown on the left and the right eye on the right. This corresponds to the visual field of the observer, so that the physiological blindspots are directed temporally (toward the temple, away from the nose) for both eyes. This is in contrast to presentation of structural data, which is usually shown like it would look in a patient sitting in front of you, where his right eye is on the observers left side.

Note that you have to specify what type of data is contained in the visual field tables. The default is set to \textbf{total deviation}, `td`.

@@

In this graph, yellow represents probabilities < 0.05, orange probabilities < 0.02, red < 0.01, and dark red < 0.005.

A gallery with all available visual field plots is available online at \url{
https://rpubs.com/huchzi/1307927}.

\subsection{Longitudinal analyses}
Monitoring visual function over time in glaucoma patients is an essential clinical application of visual fields in ophthalmology and optometry practices.

For several reasons, measurements will fluctuate over time. Fluctuation between measurements will be greater in eyes with glaucoma, especially in areas of glaucoma-related visual field defects. Therefore, it is not possible to compare two field tests and claim that the disease has progressed only because the second test yielded lower sensitivities.

Statistical methods and a series with a minimum number of tests are necessary in order to demonstrate disease progression in the visual fields. Statistical methods for progression in glaucoma that can be subdivided into \textbf{event-based} and \textbf{trend-based} methods. Furthermore, methods can be subdivided into those that require datasets which describe the "normal fluctuation" in a given population, those that rely on a mathematical model of visual field progression, and those that do not rely on either.

Event-based methods are not implemented in the visualFields package. Briefly, they look for progression between the most recent field and a baseline. Progression is characterized by a decrease in sensitivity with a magnitude that is unlikely (< 5\%) in the population under examination (population based data is necessary) and needs to be confirmed in further tests (usually at least two).

Trend-based methods assume that progression takes place more slowly, so that several field tests are carried out during this time span. This concept is unsettling for the clinician who wants to stop progression the first time that he sees the patient, but regularly takes place even in the case of optimal treatment. The advantage is that all measurements are taken into account, not only the baseline test and the most recent tests.

There are several tools to analyze progression with visualFields. To review them, we will select a series of 27 visual fields that were obtained over several years.

@@

The plot used for showing progression are shown in the gallery, which can be found at: \url{https://rpubs.com/huchzi/1307927}.

\subsection{Statistical analysis}
Statistical analysis can be performed with linear regression of global indices or with pointwise linear regression (one regression line for each location), which results in as many slopes and $p$-values as there are visual field locations (52 for the 24-2 grid after removing the two that fall on the anatomical region of the blind spot).

The \textbf{glr} function performs linear regression on global parameters. In the following example, the patient loses `r round(abs(glr(getgl(example3))$sl), 2)` dB each year and this change is statistically significant, with a $p$-value < `r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)`.

# md_regression <- data.frame(time = glr(getgl(example3))$years,
#            md = glr(getgl(example3))$data)
# 
# globallm <- glr(getgl(example3))
# slope <- globallm$sl
# intercept <- globallm$int
# pval <- globallm$pval
# 
# ggplot(md_regression, aes(x = time, y = md)) +
#   geom_point() +
#   geom_abline(slope = slope, intercept = intercept)
# 
# print(paste("intercept:", intercept))
# print(paste("slope:", slope))
# print(paste("p-value:", pval))

@@

The `plr` function performs pointwise linear regression (i.e. linear regression for each location in the field separately). Specific parameters have to be chosen, in order to detect disease progression. For example, progression might be defined as at least 3 locations having a slope less than 0 dB / year and a $p$-value < 0.05.

prog1 <- plr(example3)
str(prog1, max.level = 1)

# show the total number of siginficantly progressions locations
# (out of the 52 pointwise linear regressions)
print(sum(prog1$pval[-getlocmap()$bs] < 0.05))

The results of `plr` can be displayed using the following:

vfplotplr(example3)

@
In this graphical display, yellow means that the $p$-value for this location is < 0.05, orange that it is < 0.02, red < 0.01, and dark red < 0.005.

Criteria such as the one suggested earlier based on clusters of locations significantly progressing are useful clinically but do not control for specificity. An alternative method that ensure that specificity of progression is mantained at 95\% or any other desired level is Permutation of Pointwise Linear Regression or PoPLR [@IMF:Ole2012] [@IMF:Marin2021]. For this method, all 52 $p$-values are combined into a single $S$-statistic, originally introduced by Fisher [@IMF:Fisher1925]. In order to avoid reliance on population statistics or parametric statistics. (Fisher's original work showed that such statistic followed a chi-square distribution but only if the 52 significance tests were independent of each other and the data was Gaussian, two assumptions that are far from reality.) Thus, random permutations of the series of visual fields are generated and, for each, the $S$-statistic computed. If enough random permutations are performed, the resulting set of $S$-values represent an empirical distribution for the null hypothesis that there is no progression. A combined $p$-value can then be obtained to test this hypothesis against deterioration of the visual field. The interpretation of the $p$-value is as with any significance test.

The \textbf{poplr} function yields a large list with many intermediate computations. The main result is the $S$-value and its $p$-value. Note that there is a left tailed (deterioration), a right tailed (improvement) and a two-tailed (visual field not stable) versions in the current implementation.

prog2 <- poplr(example3)
str(prog2, max.level = 1)

An improvement is highly unlikely in this series.

print(prog2$csr)
print(prog2$csrp)

In contrast, a deterioration is very likely (p < 0.001).

prog2$csl
prog2$cslp

In the histogram representing the null distribution, you can see that a $S$-value of `r round(prog2$csl)` was only very rarely found in the random permutations of the visual field series.

hist(prog2$cstats$cslall, xlab = "S-values", xlim = c(0, 200))
abline(v = prog2$csl, col = "red", lwd = 2)

\subsection{Putting everything together}
Single field analysis reports can be created as pdfs or as interactive reports with Shiny. To explore them, type in the R Console the following:

<<eval = FALSE>>=
vfsfa(example2, file = "test.pdf")
vfsfashiny(example2)
@

\subsection{Structure function correlations in glaucoma patients}
In the last years, techniques that provide detailed imaging of the retina and the optic nerve have made incredible advances. Therefore, correlation between structure and function is becoming more important than ever.

To aid comparisons between structural insult and functional defect, it is useful to understand and model the anatomy of a typical healthy eye, even if such modeling through is imperfect given the large inter-individual differences in anatomical features [@IMF:Jan2009] [@IMF:Jan2012] [@IMF:Drasdo2007] [@IMF:Montesano2020]

In glaucoma, the damage occurs in the optic nerve at the levels of the nerve fiber bundles, which consist of the axons of the retinal ganglion cells. In the retina, these cells form two layers: the ganglion cell layer that contains the cell bodies, and the retinal nerve fiber layer. However, the cell bodies of the retinal ganglion cells are displaced from the area where the photoreceptors are located that they receive input from. This displacement is relevant mostly in the center of the visual field, so it would greatly affect the structure-function comparisons on grids such as the 10-2. The following figure shows the visual field locations and the corresponding ganglion cell soma being stimulated.

locmap <- locmaps$p10d2
gcmap <- vf2gc(locmap$coord)

plot_grid <- function() {
  lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
  lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
  lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
  lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
  lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
  lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
  for(r in c(2.5, 5, 7.5, 10))
    draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
  text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
  text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
}

par(mfrow = c(1, 2))
plot(x = locmap$coord$x, 
     y = locmap$coord$y, 
     type = "n", 
     asp = 1,
     xlim = c(-10, 12),
     ylim = c(-10, 10),
     yaxt = 'n',
     xaxt = 'n',
     bty = 'n',
     xlab = "",
     ylab = "",
     main = "10-2 grid of test locations")
plot_grid()
points(x = locmap$coord$x, y = locmap$coord$y, bg = "white")

plot(x = locmap$coord$x, 
     y = locmap$coord$y, 
     type = "n", 
     asp = 1,
     xlim = c(-10, 12),
     ylim = c(-10, 10),
     yaxt = 'n',
     xaxt = 'n',
     bty = 'n',
     xlab = "",
     ylab = "",
     main = "Corresponding ganglion cells")
plot_grid()
points(x = gcmap$x, y = gcmap$y, pch = 21, bg = "white")
@

Therefore, the visualFields package provides function to calculate where the ganglion cells that subserve a given location in the visual field are located and to calculate at which angle the axons of these ganglion cells enter the optic nerve head (optic disc), where they pass through the sclera (wall of the eye) to form the optic nerve. This function is based on Montesano et al. [@IMF:Montesano2020] which revisited the original work by Drasdo et al. [@IMF:Drasdo2007].

print(locmaps$p24d2$coord[23,])
print(vf2gc(locmaps$p24d2$coord[23,]))

The Montesano Ganglion-cell displacement function can be in conjunction with the Jansonius map describing average ganglion cell bundles paths reaching different regions of the retina [@IMF:Jan2009] [@IMF:Jan2012]. First, we can obtain the angle of incidence on the optic never head (ONH)

print(loc2psi(vf2gc(locmaps$p24d2$coord[23,])))

And from the angle of incidence, the nerve fiber bundle path can be calculated. Note that the function for calculating the trajectory will extrapolate the trajectory beyond the location of the visual field (usually to the raphe).

<<warning=FALSE, fig=TRUE>>=

# use loc2psi to calculate the incidence of the nerve fiber corresponding to a given location at the ONH
angle_of_incidence_l19 <-loc2psi(locmaps$p24d2$coord[19, ])

# use cart2jpolar for calculating the position of the visual field location in the polar coordinate system
locations <- rbind(cart2jpolar(locmaps$p24d2$coord[19, ]),
                   cart2jpolar(locmaps$p24d2$coord[23, ]),
                   cart2jpolar(locmaps$p24d2$coord[47, ]),
                   cart2jpolar(data.frame(x = 0, y = 0)))
locations$labels <- c("l19", "l23", "l47", "fixation")

# use bundePath to create a function for calculating the bundle path
path_function_l19 <- bundlePath(angle_of_incidence_l19)
path_function_l23 <- bundlePath(loc2psi(locmaps$p24d2$coord[23, ]))
path_function_l47 <- bundlePath(loc2psi(locmaps$p24d2$coord[47, ]))

bundle_path <- data.frame(y = 4:90,
                          x_l19 = path_function_l19(4:90),
                          x_l23 = path_function_l23(4:90),
                          x_l47 = path_function_l47(4:90))
ggplot(bundle_path, aes(y = y)) +
  geom_line(aes(x = x_l19)) +
  geom_point(data = locations, aes(x = -psi, y = r, color = labels, shape = labels == "fixation")) +
  geom_line(aes(x = x_l23), color = "red") +
  geom_line(aes(x = x_l47), color = "blue") +
  coord_polar(start = pi / 2, direction = -1) +
  scale_x_continuous(breaks = c(-180, -90, 0, 90, 180), limits = c(-180, 180)) +
  scale_y_continuous(limits = c(0, 60)) +
  scale_color_manual(values = c("l19" = "black", "l23" = "red", "l47" = "blue", "fixation" = "black")) +
  guides(color = "none", shape = "none")

In this plot, the origin corresponds to the center of the ONH, the black triangle corresponds to the fovea, and the $180^\circ$ angle corresponds to the line between the ONH center and the fovea, which is, in reality, slightly tilted downward in most patients. I DO NOT SEE THIS, BUT I MAY BE MISSING SOMETHING. ALSO THE FOVEA (TRIANGLE IF I GOT IT RIGHT?) SHOULD BE A BIT BELOW THE HORIZONTAL MERIDIAN. THE POSITION OF THE CENTER OF THE BLIND SPOT IS (15,+2) IN THE MODEL (PLUS AND NOT MINUS BECAUSE THE MODEL IS SPECIFIED FOR THE ANATOMICAL RETINA, AND IT NEEDS TO BE TURNED UPSIDE DOWN TO MATCH THE VISUAL FIELD)

>>>>

The \pkg{visualFields} package is available from CRAN under the Apache 2.0 license and can be installed with:

<<eval=FALSE>>=
install.packages("visualFields")
@

The latest development version can be installed from GitHub using:

<<eval=FALSE>>=
library(devtools)
install_github("imarinfr/vf1/source")
@

\subsection{Grid patterns}

Test locations are defined by location maps (\texttt{locmap}), which specify the spatial coordinates of stimulus presentation points in degrees of visual angle. These maps reflect the layout used by visual field testing devices. When analyzing data, the correct \texttt{locmap} must be specified.

By default, the \texttt{24-2} pattern is used, consisting of 54 regularly spaced locations across the central visual field.

<<>>=
library(visualFields)
default_locmap <- getlocmap()
str(default_locmap)
@

Each \texttt{locmap} contains:
\begin{itemize}
  \item A coordinate table with $x$- and $y$-positions (in degrees),
  \item A vector \texttt{bs} indicating locations corresponding to the physiological blind spot.
\end{itemize}

Other patterns are available in the package, including the \texttt{10-2} grid:

@@

All available patterns can be listed with:

@@

If the blind spot is outside the tested region (e.g., in the 10$^\circ$ pattern) or is skipped (e.g., in the Octopus G1), the \texttt{bs} field is set to \texttt{numeric(0)}.

@@

\subsection{Loading data}

The package includes example data from a glaucoma patient (\texttt{vfpwgSunyiu24d2}), consisting of 42 visual field tests.

<<>>=
str(vfpwgSunyiu24d2[, 1:10])
@

Each row represents a monocular test. The eye is labeled using Latin abbreviations: OD (right eye), OS (left eye), and OU (both eyes). Initial columns store patient information and reliability metrics. Actual test values begin at the column number stored in the \texttt{locini} variable:

<<>>=
getlocini()
getvfcols()
vfpwgSunyiu24d2[1, getvfcols()]
@

No columns should appear after these tsest values. Use \texttt{vffilter} to select specific tests and \texttt{vfisvalid} to verify data structure:

<<>>=
example1 <- vffilter(vfpwgSunyiu24d2, id == "sample1" & eye == "OD")[1, ]
example1$newVariable <- "invalid"
vfisvalid(example1)
@

\subsection{Statistical analysis of contrast thresholds}

\subsubsection{Defining defective locations}

Sensitivity is measured in decibels (dB), derived from contrast thresholds. Sensitivity varies:
\begin{itemize}
  \item Intra-individually (test-retest variability),
  \item Inter-individually (e.g., with age),
  \item By eccentricity.
\end{itemize}

To detect abnormal points, we compare observed sensitivities with an age-corrected mean normal field. These comparisons require a \textbf{normative dataset} and location-wise \textbf{age-linear models}.

The difference between observed and normal values is the \textbf{total deviation (TD)}. In this package, negative values indicate lower-than-normal sensitivity, following the Humphrey Field Analyzer convention. For perimeters that reverse this definition (e.g., Octopus), users must invert the sign.

To reveal focal defects masked by generalized depression (e.g., due to cataract), a \textbf{pattern deviation (PD)} is computed by subtracting a general height (GH) estimate from the TD. For 24-2, GH is defined as the 7th highest TD value.

\subsubsection{Normative datasets}

The package includes three normative datasets:

\begin{itemize}
  \item SUNY-IU dataset for 24-2 with Goldmann size III \citep{IMF:Mar2013b}
  \item Iowa PC26 and Peri datasets for Goldmann size V \citep{IMF:Marin2018c}
\end{itemize}

<<>>=
str(vfctrSunyiu24d2[, 1:10])
str(vfctrIowaPC26[, 1:10])
str(vfctrIowaPeri[, 1:10])
@

To build the normative model:

<<>>=
nv <- nvgenerate(vfctrSunyiu24d2)
setnv(nv)
@

The normative object \texttt{nv} includes coefficients of age-linear models:

<<>>=
print(nv$agem$coeff)
@

The expected field at age 50 is:

<<>>=
age <- 50
print(nv$agem$model(age))
@

Blind spot locations are returned as \texttt{NA}.

\subsubsection{Total deviation}

Total deviation is the difference between observed and expected sensitivity:

<<>>=
example2 <- vffilter(vfpwgRetest24d2, id == "1", eye == "OD")[1, ]
age <- example2$age[1]
normal_values <- getnv()$agem$model(age)[1:10]
example2[, getvfcols()[1:10]] - normal_values
@

Or, use the automated function:

<<>>=
td <- gettd(example2)
print(td[, getvfcols()[1:10]])
@

\subsubsection{Pattern deviation}

Pattern deviation is obtained by subtracting the general height:

<<>>=
gh <- getgh(td)
pattern_defects <- td[, getvfcols()[1:10]] - gh
@

Or automatically:

<<>>=
pd <- getpd(td)
print(pd[, getvfcols()[1:10]])
@

\subsubsection{Probability values}

Even healthy eyes may show small deviations. To interpret these values, we calculate the probability that a deviation reflects pathology using empirical distributions from the normative data \citep{IMF:Hei1987a, IMF:Heijl1991}.

<<>>=
print(getnv()$lut$probs)
print(gettdp(td)[, getvfcols()[1:10]])
print(getpdp(pd)[, getvfcols()[1:10]])
@

Probabilities less than 0.05 suggest the location is likely abnormal.

\subsection{Global indices}

Global indices summarize the visual field:

<<>>=
gl <- getgl(example2)
print(gl[, -(1:getlocini() - 1)])
@

\begin{itemize}
  \item \texttt{msens}: Mean sensitivity (MS)
  \item \texttt{ssens}: SD of sensitivity
  \item \texttt{tmd}: Mean total deviation (MD)
  \item \texttt{tsd}: SD of total deviation
  \item \texttt{pmd}: Mean pattern deviation
  \item \texttt{psd}: SD of pattern deviation
  \item \texttt{gh}: General height
  \item \texttt{vfi}: Visual field index (VFI) \citep{IMF:Ben2008}
\end{itemize}

Associated probabilities:

<<>>=
print(getglp(gl)[, -(1:getlocini() - 1)])
@

@@

\subsection{Plotting the visual field}

There are various ways to visualize perimetric data. These visualizations should be interpreted as simplified representations of the hill of vision.

Standard displays include color-coded probability maps, graphical representations of the blind spot, and border enhancements that encode diagnostic information. For example, 	exttt{vfplot} with the argument 	exttt{"td"} generates plots that require normative data on differential light sensitivity and the probabilities of deviations from these norms. This information is included for the Humphrey Field Analyzer patterns but must be provided separately for other perimeters.

By convention, when displaying both eyes side-by-side, the left eye appears on the left and the right eye on the right. This reflects the patient's visual fields, with blind spots positioned temporally. This is in contrast to structural imaging, which is typically displayed as viewed by the examiner.

It is important to specify the type of data represented in the visual field tables. The default is set to \textbf{total deviation} (	exttt{td}).

@@

In this plot, yellow indicates a probability $<$ 0.05, orange $<$ 0.02, red $<$ 0.01, and dark red $<$ 0.005.

A gallery of all available plot types is available at \url{https://rpubs.com/huchzi/1307927}.

\subsection{Longitudinal analyses}

Monitoring visual function over time is a key application of perimetry in managing glaucoma.

Sensitivity measurements naturally fluctuate between tests. This variability increases in eyes with glaucoma, especially at affected locations. Hence, two tests showing differences cannot alone indicate progression. Instead, statistical analysis across a series of tests is required to detect significant change.

Progression analysis methods fall into two broad categories: \textbf{event-based} and \textbf{trend-based} approaches. These may further be classified by their reliance on population-based normative data, mathematical models of progression, or model-free methods.

Event-based methods (not implemented in \pkg{visualFields}) compare a test to a baseline and flag change if the sensitivity reduction is unlikely ($p < 0.05$) based on normative variability, and if confirmed in subsequent tests.

Trend-based methods assume slower change and analyze all tests in a series. Though less intuitive for clinicians, these methods are statistically robust and reflect real-world progression patterns.

To illustrate, we select a time series of 27 visual fields from one eye:

@@

Example plots used for monitoring progression are available in the gallery: \url{https://rpubs.com/huchzi/1307927}.

\subsection{Statistical analysis}

Statistical progression analysis may be performed via:
\begin{itemize}
\item \textbf{Global linear regression} (e.g., on MD)
\item \textbf{Pointwise linear regression} (PLR)
\end{itemize}

The \texttt{glr()} function estimates a regression model for global indices. For the example series, the patient loses \texttt{r round(abs(glr(getgl(example3))$sl), 2)} dB/year, with $p < \texttt{r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)}$.

Pointwise linear regression is performed using \texttt{plr()}, resulting in one slope and $p$-value per location:

<<>>=
prog1 <- plr(example3)
str(prog1, max.level = 1)
print(sum(prog1$pval[-getlocmap()$bs] < 0.05))
@

To visualize these results:

<<fig=TRUE>>=
vfplotplr(example3)
@

In this display, yellow indicates $p < 0.05$, orange $< 0.02$, red $< 0.01$, and dark red $< 0.005$.

PLR-based criteria (e.g., 3 or more progressing points) are clinically useful but lack explicit control for specificity. \textbf{Permutation of Pointwise Linear Regression (PoPLR)} addresses this by combining the 52 $p$-values into a single $S$-statistic (per Fisher's method) and comparing it to a null distribution from permuted series [@IMF:Ole2012; @IMF:Marin2021].

<<>>=
prog2 <- poplr(example3)
str(prog2, max.level = 1)
@

The null distribution and observed statistic:

<<>>=
print(prog2$csr)
print(prog2$csrp)
print(prog2$csl)
print(prog2$cslp)
@

Visualization:

<<>>=
hist(prog2$cstats$cslall, xlab = "S-values", xlim = c(0, 200))
abline(v = prog2$csl, col = "red", lwd = 2)
@

\subsection{Single field reports}

Static reports and Shiny applications can be generated:

<<eval=FALSE>>=
vfsfa(example2, file = "test.pdf")
vfsfashiny(example2)
@

\subsection{Structure-function correlation}

Anatomical modeling aids comparison between structural imaging and functional testing in glaucoma.

The ganglion cell bodies stimulated by a visual field location are displaced from the photoreceptor input region, especially in the central field. This displacement is modeled in \pkg{visualFields} based on Montesano et al. [@IMF:Montesano2020].

<<>>=
locmap <- locmaps$p10d2
gcmap <- vf2gc(locmap$coord)
@

A plot of the 10-2 grid and corresponding ganglion cell bodies:

plotting code from original omitted for brevity

@@

Nerve fiber paths from each location to the optic disc can also be modeled, using Jansonius' framework [@IMF:Jan2009; @IMF:Jan2012].

<<>>=
psi23 <- loc2psi(vf2gc(locmaps$p24d2$coord[23,]))
ang23 <- psi2oct(psi23)
print(ang23 - 360)
@

\subsection{Progression reports}

Automated reports can also be generated for progression series:

<<eval = FALSE>>=
vfspa(example3, file = "test.pdf")
vfspashiny(example3)
@

\subsection{DICOM support}

DICOM standards for static perimetry are defined in Supplement 146. The package includes preliminary functions for reading HFA-generated DICOM files, but exporting to DICOM is not yet supported.

\subsection{Support for other perimeters}

Perimetry conventions vary across manufacturers. The package includes normative data for Humphrey 24-2 (Goldmann III) and Iowa full-field perimetry with Goldmann V [@IMF:Marin2018c]. The \texttt{loadoctopus} function provides partial support for Octopus G1 fields, but normative support is incomplete.

Users may provide custom normative datasets using \texttt{nvgenerate()} and \texttt{setnv()}.

\subsection{Setting the environment}

Each analysis requires environment variables:
\begin{itemize}
\item \texttt{nv}: normative values
\item \texttt{locmap}: test grid
\item \texttt{gpar}: graphical parameters
\item \texttt{locini}: first sensitivity column
\end{itemize}

<<>>=
print(names(locmaps))
print(names(normvals))
print(names(gpars))
@

These must be consistent. For example:

<<>>=
setlocmap(locmaps$pPC26)
setnv(normvals$iowa_PC26_pw)
setgpar(gpars$pPC26)
setlocini(24)
@

To reset to default (Humphrey 24-2):
